Open Policing Project Tutorial

### 2019-03-12 tutorial.Rmd
Fix typos and standardize formatting.
Also, make sure most lines are around 80 characters.

To do this we'll want to

1. use the `filter()` function to specify the years we want, and 
2. use the assignment operator `<-` (it's like = in `R`) to overwrite `stops`.

@@

The `filter()` function is used to separate rows from the dataframe that
interest us from rows that do not. In particular, `filter(DATA, CONDITION)`
returns `DATA` with all of the rows that don't satisfy `CONDITION` removed. For
instance, we might want to only look at stops where the subject was arrested. To
do this, we would type `filter(stops, arrest_made == TRUE)`, since we only want
rows from `stops` where the `arrest_made` column is (i.e., `==`) `TRUE`. (Note:
we could actually just type `filter(stops, arrest_made)`. Do you see why?)

@@

Let's make another table that gives us the proprtion of stops by race.
(There are a few equivalent ways to do this -- choose the method that feels
most natural to you.)

@@

We used ``ggplot`` to make this plot. This package provides a lot of options for
customization, and there are many tweaks we could make to this plot to make it
nicer. Its syntax is a little complicated, though, and in our data
exploration, it's enough just to get a coarse visual like this.

@@

**Fun fact:** This same trend does _not_ hold true for pedestrian stops. In fact,
if we hadn't filtered to only vehicular stops, we wouldn't have noticed this,
because this same plot over the combined pedestrian + vehicular data looks
unremarkable. The takeaway is that looking at trends by sub-categories can often
be very helpful. (E.g., in Nashville, looking at the different listed stop 
reasons uncovers the extent of the disparities.)

@@

As an eyeball comparison leads us to see that blacks are being stopped
disproportionately, relative to the city's population. But let's be a bit more
rigorous about this. If we join the two tables together, we can compute stop 
rates by race (i.e., number of stops per person). Remember to take into acount
how many years are in your stop data, in order to get a true value of stops per
capita; we're using only 2017 for stops and for population, so we're in good 
shape.

@@

The way we put tables together is with the `left_join()` function. We need to
input three things into this function: the two tables we'd like to combine,
along with instructions on how to join them. In this case, the two tables we
want to join is the table of stops counted according to `subject race` and
`population_2017`. The instruction for combining the tables is to join rows that
contain the same race, so we have to also give `left_join()` the argument `by =
"subject_race"`. This means that `left_join()` will look at the first table,
i.e., the table stops counted by race, and then go to the `subject_race` column
in each row. Then, `left_join()` will take what it finds there---in this case,
`"asian/pacific islander"`, `"black"`, `"hispanic"`, `"other/unknown"`, and
`"white"`---and look in the second table, i.e., `population_2017`, for all the
rows that contain the same information in `population_2017`'s race column. Then,
it will add the second row at the end of the first to create a new row with
information from both. What we end up with is a dataframe with all of the
columns from __both__ tables.

@@

Good! Now we can divide the black stop rate by the white stop rate to be able to make 
a quantitative statement about how much more often blacks are stops versus whites,
relative to their share of the city's population. Blacks are stopped at a rate 3.4 
times higher than whites are, and Hispanics are stopped at a rate 1.5 times higher 
than whites are.

@@

The second step is to do something with those groups. That's what `summarize()`
does. The way `summarize()` works is, given a dataframe broken into groups by
`group_by()`, will calculate a value, one for each group. The basic syntax is
`summarize(DATA, STAT = FUN(COL_NAME))`, where `DATA` is some dataframe broken
up by `group_by()`, `STAT` is some statistic we want to calculate, `FUN` is the
function that calculates that statistic, and `COL_NAME` is the name of the
column (or columns) used to calculate the statistic.

@@

Note that with functions like `mean()`, if any of the values that are being
averaged are `NA`, the output value of the mean will simply be `NA`. We don't have
to worry about that with our Philly data, because it's super clean. But you may
encounter this problem in some of your own data. The easy workaround is to pass
the argument `na.rm = T` into `mean()` -- as you see in the code above. This tells 
`R` to remove the `NA` values and compute the mean over all the non-NA values.

Now let's dive into these results! As with the stop rates, we can make a 
quantitative claim about disparities in
search and frisk rates by dividing the minority rate by the white rate.
Here we see that among drivers who were stopped, blacks were searched at a rate
1.5 times higher than whites were, and Hispanics were searched at a rate 1.2
times higher than whites were. Black drivers were frisked at a rate 2.1 times
higher than white drivers were, and Hispanic drivers were frisked at a rate
1.5 times higher than white drivers were.

@@

For the stop rate benchmark, what we really want to know is what the true 
distribution is for individuals breaking traffic laws and/or exhibiting other 
criminal behavior in their vehicles. If blacks and Hispanics are disproportionately 
stopped relative to their rates of offending, that would be stronger evidence.
Some people then proposed to use benchmarks that approximate those offending
rates, like arrests, for example. However, we know arrests to themselves be 
racially skewed (especially for low-level drug offenses, for example), so it 
wouldn't give us the true offending population's racial distribution. Furthermore,
only `stops %>% summarize(p = round(100 * mean(arrest_made), 1)) %>% pull(p)`%
of stops result in an arrest, so the arrested population will naturally not
match the stopped population.

An even simpler critique of the population bechmark for stop rates is that it
doesn't account for possible race-specific differences in driving behavior, including
amount of time spent on the road (and adherence to traffic laws, as mentioned above).
If black drivers, hypothetically, spend more time on the road than whites, that in and
of itself could explain the higher stop rates we see for black drivers, even in the
absence of discrimination.

Search and frisk rates are slightly less suspect, since among the stopped 
population, it's more reasonable to believe that people of different races
offend at equal rates. In the context of searches, this means assuming that
all races exhibit probable cause of possessing contraband at equal rates. And
in the case of frisks, this means assuming that all races exhibit reasonable
articulable suspicion of possessing a weapon at equal rates. Of course, these
assumptions can also be argued against. One could claim that the stopped population
isn't a good measure of the true racial distribution of probable cause.

This is all to say that while benchmark stats are a good place to start, more
investigation is required before we can draw any conclusions.

@@

We thus turn to rates of successful searches. That is, what proportion of
searches, by race, were successful? This proportion is known as the contraband
recovery rate, or the "hit rate". If racial groups have different hit rates, it
can imply that racial groups are being subjected to different standards.

@@

We see that hit rates are slightly lower for blacks and Hispanics than for
whites.

@@

Again, this is too many hit rates to compare in one table. To plot the hit rates
of blacks vs. whites and of Hispanics vs. whites, we need to reshape our table
to have each row containing a district, a minority race, minority hit rate in
that district, and white hit rate in that district. We'll walk you through the 
code below that reshapes the data for us.

```{r}
# Reshape table to show hit rates of minorities vs whites
hit_rates <-
  hit_rates %>% 
  filter(subject_race %in% c("black", "white", "hispanic")) %>% 
  spread(subject_race, hit_rate, fill = 0) %>% 
  rename(white_hit_rate = white) %>% 
  gather(minority_race, minority_hit_rate, c(black, hispanic)) %>%
  arrange(district)

@@

**Quick Tip**: You might have noticed a strange symbol in the code above:
`%in%`. This is really just shorthand for the sentence, "the `subject_race` is
`"black"` or it is `"white"` or it is `"hispanic"`." Without the `%in%` syntax, we
would have to write this in `R` as: `subject_race == "black" | subject_race ==
"white" | subject_race == "hispanic"`. The `%in%` syntax is a lot shorter and
clearer; we think you'll agree.

@@

hit_rates %>% 
  ggplot(aes(
    x = white_hit_rate,
    y = minority_hit_rate
  )) +
  geom_point() +
  # This sets a diagonal reference line (line of equal hit rates)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  # These next few lines just make the axes pretty and even
  scale_x_continuous("White hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  scale_y_continuous("Minority hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  # This makes sure that 1% on the x-axis is the same as 1% on the y-axis
  coord_fixed() +
  # This allows us to compare black v. white and Hispanic v. white side by
  # side, in panels
  facet_grid(. ~ minority_race)
  # Depending on your version of ggplot2, you may be able to use the syntax 
  # below (the newer ggplot2 syntax) -- which is clearer, in my opinion.
  # But older versions of ggplot2 will only accept the above syntax
  # facet_grid(cols = vars(minority_race))
```

@@

It seems like we can now say with confidence that even looking within location,
nearly every police district has lower hit rates for minorities than whites.

@@

Depending on how you investigated the `location`, you might have found that the
majority of the stops happen at the airport, or have location listed as NA or
unavailable. Indeed, if we check the Philadelphia Police's website, we'll find
that there is no District 77, and that the airport does not fall within the 
jurisdiction of any particular police district. Between this information and the
fact that hit rates are essentially zero for this district, we can conclude that
the types of searches happening in this location are qualitatively different
from searches in our other locations.

@@

Wow! A sizeable share of searches of white individuals must have been taking place
at the airport with zero hit rates, deflating the white hit rate we were 
calculating before. But this didn't influence hit rates of blacks or Hispanics
as much. Our aggregate hit rate analysis now shows stronger evidence of bias,
similar to our by-location plot. (Note, thought, that aggregate and by-location 
hit rates do not necessarily always tell the same story -- a phenomenon known as
Simpson's Paradox -- which is why it's important to check both to get the full 
picture.)

@@

(**Fun fact:** The threshold test confirms the results we see in Philadelphia.
Officers are indeed applying lower thresholds when deciding to search black and 
Hispanic drivers than when deciding to search white drivers.)

@@

The outcome test is a good way to get around the benchmarking problem at the
search decision, but when it comes to assessing bias at the stop decision,
it's a little trickier. The outcome test usually doesn't work, since rarely
is there a notion of what a "successful" stop is. (There are, of course, some
exceptions to this, like when stop reason is listed as suspected Criminal 
Possession of a Weapon (CPW) -- which is the case in NYC stop-question-frisk
data.)

@@

To adjust for clock time, we want to compare times that were light at some
point during the year (i.e., occurred before dusk) but were dark at other points
during there (i.e., occurred after dusk). This is called the "inter-twilight
period": the range from the earliest time dusk occurs in the year to the latest 
time dusk occurs in the year.

@@

So the earliest sunset in 2017 was at around 4:30pm, in December (and it was
fully dark about 30 minutes later, at dusk, around 5pm). The latest sunset time
in 2017 was at around 8:30pm, in late June (and it was fully dark about 30 
minutes later, at dusk, around 9pm).

For a time in this range, like 6:30pm, part of the year it's still light, and part
of the year, it's still dark. So we can compare the proportion of black drivers 
stopped at 6:30pm when it's light (usually in the summer) to the proportion of 
black drivers stopped at 6:30pm when it's dark (usually in the winter). This way,
we're really seeing the effect of darkness, since driving and deployment 
patterns tend to change only with clock time. For example, the demographics
of commuters who are on the road at 6:30pm likely stays pretty steady throughout
the year, compared to demographics of commuters at 5pm versus at 9pm.

One quick note: We want to make sure to filter out stops that occurred in the
approximately 30-minute period between sunset and dusk (end of civil twilight),
since it is ambiguous as to whether that period is light or dark, which would
muddle up our analysis.

Let's join the sunset times to our stops data and

* filter to the inter-twilight period
* remove that ambigous pre-dusk period
* for simplicity, compare only black and white drivers

```{r}
vod_stops <- 
  stops %>% 
  left_join(
    sunset_times,
    by = "date"
  ) %>% 
  mutate(
    minute = time_to_minute(time),
    minutes_after_dark = minute - dusk_minute,
    is_dark = minute > dusk_minute,
    min_dusk_minute = min(dusk_minute),
    max_dusk_minute = max(dusk_minute),
    is_black = subject_race == "black"
  ) %>% 
  filter(
    # Filter to get only the intertwilight period
    minute >= min_dusk_minute,
    minute <= max_dusk_minute,
    # Remove ambigous period between sunset and dusk
    !(minute > sunset_minute & minute < dusk_minute),
    # Compare only whites and blacks
    subject_race %in% c("black", "white")
  )
```

How many stops are in our this new, filtered data frame?

@@

Now, to get some intuition, let's use our 6:30pm example. If we filter the data
to stops that occurred in the narrow window from 6:30pm to 6:45pm, we can compute
what proportion of stops were of black drivers for stops when it was dark and 
compare that to the proportion for stops when it was not dark yet.

@@

Indeed we see that the proportion is smaller when 6:30-6:45pm is dark than when
it's light. According to the veil of darkness hypothesis, this is because 
officers engaged in racial profiling can't determine drivers' races as well
at night.  We know that the proportion of black drivers stopped dropped by 
about 5% when it was dark and race was harder to identify. It's hard to say what
this means in terms of the extent or pervasiveness, though. We **cannot** make
*any* claims about whether some small percent of the officers were engaged in 
stops that were 100% profiled on race, or whether all officers were engaged in
racial profiling in some small percentage of their stops, or anything of the
sort.

Furthermore, while this stat gives us some intuition that racial profiling might 
be present in Philly, we've only looked at one time. In order to make sure the
result is robust, we want to check this for _every_ time in our inter-twilight
period. To do this, we can use _logistic regression_. The basic idea here is
that we're finding what our data implied about how darkness and clock time
influence the whether a stopped driver will be black. We then extract the 
coefficient of darkness (you can think of this loosely as the measure of how
much darkness -- as opposed to clock time -- influences whether a stopped driver
will be black).

@@

The fact that the coefficient is negative means that darkness _lessens_ the 
likelihood that a stopped driver will be black, after adjusting for clock 
time. This matches our intuitive result from 6:30pm.

The fact that the standard error is small means that this is a _statistically significant_ 
finding. Another important thing to do, besides checking 
statistical significance of your results, is to do a bunch of different 
robustness checks. The idea is that all models make some choices about how to 
turn the question at hand into math, and we want to vary those choices to make
sure that none of them change the model result -- that the result holds regardless
of, for example, how you choose to control for time. In our model we used a
natural spline with six degrees of freedom, but we could have chosen to use
3 degrees of freedom, or to do something way simpler like bin time to one-minute 
or five-minute windows.

Another robustness check that often comes up with this type of modeling is
adding in a control for sub-geography. For the outcome test, we controlled
for police district under the proposition that maybe it was an intentional
policy decision for certain police districts to lower search thresholds 
than other police districts -- and if those districts happened to be 
disproportionately black, that would "explain away" the bias we saw in the 
aggregate outcome test results.

In the case of the Veil of Darkness test, one could make an analogous claim:
Is the result that we're seeing just a product of certain districts being policed
more after dark and those districts have a lower proportion of black drivers 
than the districts being policed before dark? If so, then adjusting for location
in our model would lead to the darkness coefficient being zero. But if the
coefficient on darkness doesn't change much (i.e., stays negative in a 
statistically significant way), then that's another marker that our finding is
robust.

@@

The veil-of-darkness test is a popular technique for assessing disparate treatment, 
but, like all statistical methods, it comes with caveats. (Hopefully that's one of
the takeaways today -- data and statistical analysis can be very powerful, but it's
good to know what the limitations are.)
Perhaps most importantly, darkness--after adjusting for time of day--is a 
function of the date. 
As such, to the extent that driver behavior changes throughout the year, and these 
changes are correlated with race, the test can suggest discrimination where there is 
none. 
One way to account for seasonal effects is to consider the brief period around daylight
savings. Driving behavior likely doesn't change much at, say, 6pm on the day before
daylight savings to the day after; but the sky is light on one day and dark on the next.
Clever checks like this can help circumvent some of the concerns about the veil of 
darkness test. However, if driving behavior is more related to lighting than time of day, the daylight savings test wouldn't help.

@@

Other things, like vehicle make, year, and model often correlate with race and are
still visible at night, which could lead to the test under-estimating the extent of
racial profiling. Similarly, the test doesn't control for stop reason, which is often
correlated with both race and time of day (consider, for example, pretextual broken 
tail light stops).

@@

## Parting shots

Hopefully this gets you started on the basics of how to analyze data, and gives
you a little glimpse of some more complex modeling you can do with the data too.
While these analyses (from simple hit rates to modeling racial
profiling) can be really helpful in quantifying racial disparities, it's often
basic exploration of the data that yields the most interesting stories. So, the 
number one rule in working with data is **be curious!**

>>>>

To do this we'll want to:

1. Use the `filter()` function to specify the years we want, and 
2. Use the assignment operator `<-` (it's like = in `R`) to overwrite `stops`.

@@

The `filter()` function is used to separate rows from the dataframe that
interest us from rows that do not. In particular, `filter(DATA, CONDITION)`
returns `DATA` with all of the rows that don't satisfy `CONDITION` removed. For
instance, we might want to only look at stops where the subject was arrested. To
do this, we would type `filter(stops, arrest_made == TRUE)`, since we only want
rows from `stops` where the `arrest_made` column is (i.e., `==`) `TRUE`. (**Note**:
we could actually just type `filter(stops, arrest_made)`. Do you see why?)

@@

Let's make another table that gives us the proprtion of stops by race.
(There are a few equivalent ways to do this---choose the method that feels
most natural to you.)

@@

We used `ggplot` to make this plot. This package provides a lot of options for
customization, and there are many tweaks we could make to this plot to make it
nicer. Its syntax is a little complicated, though, and in our data
exploration, it's enough just to get a coarse visual like this.

@@

**Fun fact**: This same trend does _not_ hold true for pedestrian stops. In fact,
if we hadn't filtered to only vehicular stops, we wouldn't have noticed this,
because this same plot over the combined pedestrian and vehicular data looks
unremarkable. The takeaway is that looking at trends by sub-categories can often
be very helpful. (E.g., in Nashville, looking at the different listed stop 
reasons uncovers the extent of the disparities.)

@@

As an eyeball comparison leads us to see that black drivers are being stopped
disproportionately, relative to the city's population. But let's be a bit more
rigorous about this. If we join the two tables together, we can compute stop 
rates by race (i.e., number of stops per person). Remember to take into acount
how many years are in your stop data, in order to get a true value of stops per
capita; we're using only 2017 for stops and for population, so we're in good 
shape.

@@

The way we put tables together is with the `left_join()` function. We need to
input three things into this function: the two tables we'd like to combine,
along with instructions on how to join them. In this case, the two tables we
want to join are the table of stops counted according to `subject race` and
`population_2017`. The instruction for combining the tables is to join rows that
contain the same race, so we have to also give `left_join()` the argument `by =
"subject_race"`. This means that `left_join()` will look at the first table---
i.e., the table stops counted by race---and then go to the `subject_race` column
in each row. Then, `left_join()` will take what it finds there---in this case,
`"asian/pacific islander"`, `"black"`, `"hispanic"`, `"other/unknown"`, and
`"white"`---and look in the second table, i.e., `population_2017`, for all the
rows that contain the same information in `population_2017`'s race column. Then,
it will add the second row at the end of the first to create a new row with
information from both. What we end up with is a dataframe with all of the
columns from _both_ tables.

@@

Good! Now we can divide the black stop rate by the white stop rate to be able to make 
a quantitative statement about how much more often black drivers are stops
compared to white drivers, relative to their share of the city's population.
Black drivers are stopped at a rate 3.4 times higher than white drivers, and
Hispanic drivers are stopped at a rate 1.5 times higher than white drivers.

@@

The second step is to do something with those groups. That's what `summarize()`
does. The way `summarize()` works is to take a dataframe broken into groups by
`group_by()` and calculate a statistic for each group . The basic syntax is
`summarize(DATA, STAT = FUN(COL_NAME))`, where `DATA` is some dataframe broken
up by `group_by()`, `STAT` is some statistic we want to calculate, `FUN` is the
function that calculates that statistic, and `COL_NAME` is the name of the
column (or columns) used to calculate the statistic.

@@

Note that with functions like `mean()`, if any of the values that are being
averaged are `NA`, the output value of the mean will simply be `NA`. We don't have
to worry about that with our Philly data, because it's super clean. But you may
encounter this problem in some of your own data. The easy workaround is to pass
the argument `na.rm = T` into `mean()`---as you see in the code above. This tells 
`R` to remove the `NA` values and compute the mean over all the non-NA values.

Now let's dive into these results! As with the stop rates, we can make a 
quantitative claim about disparities in search and frisk rates by dividing the
minority rate by the white rate. Here we see that among drivers who were
stopped, black drivers were searched at a rate 1.5 times higher than white
drivers, and Hispanic drivers were searched at a rate 1.2 times higher than
white drivers. Black drivers were frisked at a rate 2.1 times higher than
white drivers were, and Hispanic drivers were frisked at a rate 1.5 times
higher than white drivers were.

@@

For the stop rate benchmark, what we really want to know is what the true 
distribution is for individuals breaking traffic laws or exhibiting other 
criminal behavior in their vehicles. If black and Hispanic drivers are
disproportionately stopped relative to their rates of offending, that would be
stronger evidence. Some people then proposed to use benchmarks that approximate
those offending rates, like arrests, for example. However, we know arrests to
themselves be racially skewed (especially for low-level drug offenses, for
example), so it wouldn't give us the true offending population's racial
distribution. Furthermore, only `stops %>% summarize(p = round(100 *
mean(arrest_made), 1)) %>% pull(p)`% of stops result in an arrest, so the
arrested population will naturally not match the stopped population.

An even simpler critique of the population bechmark for stop rates is that it
doesn't account for possible race-specific differences in driving behavior,
including amount of time spent on the road (and adherence to traffic laws, as
mentioned above).  If black drivers, hypothetically, spend more time on the
road than white drivers, that in and of itself could explain the higher stop
rates we see for black drivers, even in the absence of discrimination.

Search and frisk rates are slightly less suspect, since among the stopped 
population, it's more reasonable to believe that people of different races
offend at equal rates. In the context of searches, this means assuming that
all races exhibit probable cause of possessing contraband at equal rates. And
in the case of frisks, this means assuming that all races exhibit reasonable
articulable suspicion of possessing a weapon at equal rates. Of course, one
could also argue against these assumptions. One could claim that the stopped
population isn't a good measure of the true racial distribution of probable
cause. This is all to say that while benchmark stats are a good place to
start, more investigation is required before we can draw any conclusions.

@@

We thus turn to rates of successful searches. That is, what proportion of
searches, by race, were successful? This proportion is known as the contraband
recovery rate, or the "hit rate." If racial groups have different hit rates, it
can imply that racial groups are being subjected to different standards.

@@

We see that hit rates are slightly lower for black and Hispanic drivers than
for white drivers.

@@

Again, this is too many hit rates to compare in one table. To plot the hit rates
of black vs. white drivers and of Hispanic vs. white drivers, we need to
reshape our table to have each row containing a district, a minority race,
minority hit rate in that district, and white hit rate in that district. We'll
walk you through the code below that reshapes the data for us.

```{r}
# Reshape table to show hit rates of minorities vs white drivers
hit_rates <-
  hit_rates %>% 
  filter(subject_race %in% c("black", "white", "hispanic")) %>% 
  spread(subject_race, hit_rate, fill = 0) %>% 
  rename(white_hit_rate = white) %>% 
  gather(minority_race, minority_hit_rate, c(black, hispanic)) %>%
  arrange(district)

@@

**Quick Tip**: You might have noticed a strange symbol in the code above:
`%in%`. This is really just shorthand for the sentence, "the `subject_race` is
`"black"` or it is `"white"` or it is `"hispanic"`." Without the `%in%` syntax, we
would have to write this in `R` as: `subject_race == "black" | subject_race ==
"white" | subject_race == "hispanic"`. The `%in%` syntax is a lot shorter and
clearer, as we think you'll agree.

@@

hit_rates %>% 
  ggplot(aes(
    x = white_hit_rate,
    y = minority_hit_rate
  )) +
  geom_point() +
  # This sets a diagonal reference line (line of equal hit rates)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  # These next few lines just make the axes pretty and even
  scale_x_continuous("White hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  scale_y_continuous("Minority hit rate", 
    limits = c(0, max_hit_rate + 0.01),
    labels = scales::percent
  ) +
  # This makes sure that 1% on the x-axis is the same as 1% on the y-axis
  coord_fixed() +
  # This allows us to compare black v. white and Hispanic v. white side by
  # side, in panels
  facet_grid(. ~ minority_race)
  # Depending on your version of ggplot2, you may be able to use the syntax 
  # below (the newer ggplot2 syntax)---which is clearer, in my opinion.
  # But older versions of ggplot2 will only accept the above syntax
  # facet_grid(cols = vars(minority_race))
```

@@

It seems like we can now say with confidence that even looking within location,
nearly every police district has lower hit rates for minorities than white
drivers.

@@

Depending on how you investigated the `location`, you might have found that the
majority of the stops happen at the airport, or have location listed as `NA` or
unavailable. Indeed, if we check the Philadelphia Police's website, we'll find
that there is no District 77, and that the airport does not fall within the 
jurisdiction of any particular police district. Between this information and the
fact that hit rates are essentially zero for this district, we can conclude that
the types of searches happening in this location are qualitatively different
from searches in our other locations.

@@

Wow! A sizeable share of searches of white individuals must have been taking
place at the airport with zero hit rates, deflating the white hit rate we
were calculating before. But this didn't influence hit rates of black or
Hispanic drivers as much. Our aggregate hit rate analysis now shows stronger
evidence of bias, similar to our by-location plot. (Note, thought, that
aggregate and by-location hit rates do not necessarily always tell the same
story---a phenomenon known as Simpson's Paradox---which is why it's important
to check both to get the full picture.)

@@

(**Fun fact**: The threshold test confirms the results we see in Philadelphia.
Officers are indeed applying lower thresholds when deciding to search black and 
Hispanic drivers than when deciding to search white drivers.)

@@

The outcome test is a good way to get around the benchmarking problem at the
search decision, but when it comes to assessing bias at the stop decision,
it's a little trickier. The outcome test usually doesn't work, since rarely
is there a notion of what a "successful" stop is. (There are, of course, some
exceptions to this, like when stop reason is listed as suspected Criminal 
Possession of a Weapon (CPW)---which is the case in NYC stop-question-frisk
data.)

@@

To adjust for clock time, we want to compare times that were light at some
point during the year (i.e., occurred before dusk) but were dark at other points
during the year (i.e., occurred after dusk). This is called the "inter-twilight
period": the range from the earliest time dusk occurs in the year to the latest 
time dusk occurs in the year.

@@

So the earliest sunset in 2017 was at around 4:30 p.m., in December (and it was
fully dark about 30 minutes later, at dusk, around 5:00 p.m.). The latest sunset time
in 2017 was at around 8:30 p.m., in late June (and it was fully dark about 30 
minutes later, at dusk, around 9:00 p.m.).

For a time in this range, like 6:30 p.m., part of the year it's still light, and part
of the year, it's still dark. So we can compare the proportion of black drivers 
stopped at 6:30 p.m. when it's light (usually in the summer) to the proportion of 
black drivers stopped at 6:30 p.m. when it's dark (usually in the winter). This way,
we're really seeing the effect of darkness, since driving and deployment 
patterns tend to change only with clock time. For example, the demographics
of commuters who are on the road at 6:30 p.m. likely stays pretty steady throughout
the year, compared to demographics of commuters at 5:00 p.m. versus at 9:00 p.m.

One quick note: We want to make sure to filter out stops that occurred in the
approximately 30-minute period between sunset and dusk (end of civil twilight),
since it is ambiguous whether that period is light or dark, which would muddle
up our analysis.

Let's join the sunset times to our stops data and:

1. Filter to the inter-twilight period;
2. Remove that ambigous pre-dusk period; and
3. For simplicity, compare only black and white drivers.

```{r}
vod_stops <- 
  stops %>% 
  left_join(
    sunset_times,
    by = "date"
  ) %>% 
  mutate(
    minute = time_to_minute(time),
    minutes_after_dark = minute - dusk_minute,
    is_dark = minute > dusk_minute,
    min_dusk_minute = min(dusk_minute),
    max_dusk_minute = max(dusk_minute),
    is_black = subject_race == "black"
  ) %>% 
  filter(
    # Filter to get only the intertwilight period
    minute >= min_dusk_minute,
    minute <= max_dusk_minute,
    # Remove ambigous period between sunset and dusk
    !(minute > sunset_minute & minute < dusk_minute),
    # Compare only white and black drivers
    subject_race %in% c("black", "white")
  )
```

How many stops are in our new, filtered data frame?

@@

Now, to get some intuition, let's use our 6:30 p.m. example. If we filter the data
to stops that occurred in the narrow window from 6:30 p.m. to 6:45 p.m., we
can compute what proportion of stops were of black drivers for stops when it
was dark and compare that to the proportion for stops when it was not dark yet.

@@

Indeed, we see that the proportion is smaller when 6:30--6:45 p.m. is dark
than when it's light. According to the veil of darkness hypothesis, this is
because officers engaged in racial profiling can't determine drivers' races as
well at night.  We know that the proportion of black drivers stopped dropped by
about 5% when it was dark and race was harder to identify. It's hard to say what
this means in terms of the extent or pervasiveness, though. We *cannot* make
*any* claims about whether some small percent of the officers were engaged in 
stops that were 100% profiled on race, or whether all officers were engaged in
racial profiling in some small percentage of their stops, or anything of the
sort.

Furthermore, while this stat gives us some intuition that racial profiling might 
be present in Philly, we've only looked at one time. In order to make sure the
result is robust, we want to check this for _every_ time in our inter-twilight
period. To do this, we can use _logistic regression_. The basic idea here is
that we're finding what our data implied about how darkness and clock time
influence the whether a stopped driver will be black. We then extract the 
coefficient of darkness. (You can think of this loosely as the measure of how
much darkness---as opposed to clock time---influences whether a stopped driver
will be black.)

@@

The fact that the coefficient is negative means that darkness _lessens_ the 
likelihood that a stopped driver will be black, after adjusting for clock 
time. This matches our intuitive result from 6:30 p.m.

The fact that the standard error is small means that this is a _statistically significant_ 
finding. Another important thing to do, besides checking 
statistical significance of your results, is to do a bunch of different 
robustness checks. The idea is that all models make some choices about how to 
turn the question at hand into math, and we want to vary those choices to make
sure that none of them change the model result---that the result holds regardless
of, for example, how you choose to control for time. In our model we used a
natural spline with six degrees of freedom, but we could have chosen to use
3 degrees of freedom, or to do something way simpler like bin time to one-minute 
or five-minute windows.

Another robustness check that often comes up with this type of modeling is
adding in a control for sub-geography. For the outcome test, we controlled
for police district under the proposition that maybe it was an intentional
policy decision for certain police districts to lower search thresholds 
than other police districts---and if those districts happened to be 
disproportionately black, that would "explain away" the bias we saw in the 
aggregate outcome test results.

In the case of the Veil of Darkness test, one could make an analogous claim:
is the result that we're seeing just a product of certain districts being policed
more after dark and those districts have a lower proportion of black drivers 
than the districts being policed before dark? If so, then adjusting for location
in our model would lead to the darkness coefficient being zero. But if the
coefficient on darkness doesn't change much (i.e., stays negative in a 
statistically significant way), then that's another marker that our finding is
robust.

@@

The veil-of-darkness test is a popular technique for assessing disparate
treatment, but, like all statistical methods, it comes with caveats.
(Hopefully that's one of the takeaways today---data and statistical analysis
can be very powerful, but it's good to know what the limitations are.) Perhaps
most importantly, darkness---after adjusting for time of day---is a function of
the date. As such, to the extent that driver behavior changes throughout the
year, and these changes are correlated with race, the test can suggest
discrimination where there is none. One way to account for seasonal effects
is to consider the brief period around daylight savings. Driving behavior
likely doesn't change much at, say, 6:00 p.m. the day before daylight savings to
the day after; but the sky is light on one day and dark on the next. Clever
checks like this can help circumvent some of the concerns about the veil of
darkness test. However, if driving behavior is more related to lighting than
time of day, the daylight savings test wouldn't help.

@@

Other things, like vehicle make, year, and model often correlate with race and
are still visible at night, which could lead to the test under-estimating the
extent of racial profiling. Similarly, the test doesn't control for stop
reason, which is often correlated with both race and time of day. (Consider,
for example, pretextual broken tail light stops.)

@@

## Parting thoughts

Hopefully this gets you started on the basics of how to analyze data, and gives
you a little glimpse of some more complex modeling you can do with the data too.
While these analyses (from simple hit rates to modeling racial
profiling) can be really helpful in quantifying racial disparities, it's often
basic exploration of the data that yields the most interesting stories. So, the 
number one rule in working with data is: *be curious!*
